partial: An R Package for Creating Partial Dependence Plots

### 2016-08-01 manuscript/greenwell.tex
major tweak to imports

\abstract{
Complex predictive models (e.g., neural networks, random forests, and support vector machines) offer much promise in the world of data analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g. multiple linear regression). Unfortunately, it can be challenging for the uninitiated to understand the results of such models and explain them to management. Partial dependence plots (PDPs) offer a simple solution. Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be easily understood. These plots are especially useful in explaining the output from complex black box models like support vector machines or neural networks or constructing simpler parsimonious models on newly obtained data.
}

@@

Predictor importance is an important task in any supervised learning problem. However, ranking variables is only part of the story and once a subset of "important" features is identified it is often necessary to assess the relationship between them and the response. This can be done in many ways, but in machine learning it is often accomplished by constructing \dfn{partial dependence plots} (PDPs); see \citet{friedman-2000-greedy} for details. PDPs help visualize the relationship between a subset (typically 1-3) of the predictors and the response while accounting for the average effect of the other independent variables in the model. They are particularly effective with black box models like random forest or support vector machines.
%PDPs offer a low-dimensional graphical rendering of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ for any type of fitted models and are especially useful in visualizing the relationships discovered by complex machine learning algorithms such as a \dfn{random forest}.

Let $\boldsymbol{x} = \left\{x_1, x_2, \dots, x_n\right\}$ represent the predictors in a model whose prediction function is $\widehat{f}\left(\boldsymbol{x}\right)$. If we partition $\boldsymbol{x}$ into an interest set, $\boldsymbol{z}_s$, and its compliment, $\boldsymbol{z}_{c} = \boldsymbol{x} \setminus \boldsymbol{z}_s$, then the partial dependence of the response on $\boldsymbol{z}_s$ is defined as
\begin{equation}
\label{eqn:pdf}
\bar{f}_s\left(\boldsymbol{z}_s\right) = \frac{1}{n}\sum_{i = 1}^n\widehat{f}\left(\boldsymbol{z}_s,\boldsymbol{z}_{i, c}\right),
\end{equation}
where $\boldsymbol{z}_{i, c}$ $\left(i = 1, 2, \dots, n\right)$ are the values of $\boldsymbol{z}_c$ that occur in the training sample. This can be quite computationally intensive since Equation~\eqref{eqn:pdf} involves a pass over the training records for each set of combinations of $\boldsymbol{z}_s$. Fortunately, this problem is embarrassingly parallel. Let $x_1$ be the predictor variable of interest with unique values $\left\{x_{11}, x_{12}, \dots, x_{1k}\right\}$. The partial dependence of the response on $x_1$ can be constructed as follows:
\begin{enumerate}
  \item For $i \in \left\{1, 2, \dots, k\right\}$:
  \begin{enumerate}
    \item Create a copy of the training data and replace the original values of $x_1$ with the constant $x_{1i}$.
    \item Compute the vector of predicted values from the modified copy of the training data.
    \item Compute the average prediction to obtain $\bar{f}_1\left(x_{1i}\right)$.
  \end{enumerate}
  \item Plot the pairs $\left[x_{1i}, \bar{f}_1\left(x_{1i}\right)\right]$ for $i = 1, 2, \dotsc, k$.
\end{enumerate}
This algorithm can be easily extended to larger subsets of 2 or more features as well.

Limited implementations of Friedman's PDPs are available in packages \CRANpkg{randomForest} \citep{randomForest-pkg} and \CRANpkg{gbm}, among others. While the \pkg{randomForest} implementation will only allow for a single predictor, the \pkg{gbm} implementation can deal with any subset of the predictor space; however, these implementations only apply to models fit using the respective package. Partial dependence functions are not restricted to tree-based models; they can be applied to any supervised learning algorithm (e.g., neural networks). However, to our knowledge, there is no general package for constructing PDPs  in R (e.g., PDPs for a \dfn{conditional random forest} as implemented by the \code{cforest} function in the \CRANpkg{party} and \CRANpkg{partykit} packages; see \citet{party-pkg} and \citet{partykit-pkg}, respectively). The \CRANpkg{partial} \citep{partial-pkg} package tries to close this gap by offering a general framework for PDPs that can be applied to several types of fitted models.

@@

Currently, only two functions are exported by \pkg{partial}:
\begin{itemize}
  \item \code{partial}
  \item \code{plotPartial}
\end{itemize}
The \code{partial} function evaluates the partial dependence \eqref{eqn:pdf} from a fitted model over a grid of predictor values. If \code{plot = FALSE} (the default), \code{partial} returns a data frame with an additional class: \code{"partial"}. The columns are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{y} and contains the partial dependence function values. If \code{plot = TRUE}, a \code{"lattice"} object representing the PDP is returned. For more advanced plotting, the \code{plotPartial} functions will take a \code{"partial"} object and display a more customizable lattice plot. Currently supported models are described in Table~\ref{tab:models}.
\begin{table}[htbp]
  \begin{tabular}{p{4cm}ll}
    \toprule
      Model & Package & Class \\
      \midrule
      Bagged decision trees     & \CRANpkg{ipred} \citep{ipred-pkg} & \code{"classbagg"}, \code{"regbagg"} \\
                                & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"bagging"} \\
      Boosted decision trees    & \pkg{gbm}      & \code{"gbm"} \\
                                & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"boosting"} \\
      Conditional random forest & \pkg{party}    & \code{"RandomForest"} \\
                                & \pkg{partykit} & \code{"cforest"} \\
      Decision trees            & \CRANpkg{rpart} \citep{rpart-pkg} & \code{"rpart"} \\
                                & \pkg{party}    & \code{"BinaryTree"} \\
                                & \pkg{partykit} & \code{"constparty"}/\code{"party"} \\
      Linear model              & \pkg{stats}    & \code{"lm"} \\
      Multivariate adaptive regression splines (MARS) & \CRANpkg{earth} \citep{earth-pkg} & \code{"earth"} \\
      Random forest & \pkg{randomForest} & \code{"randomForest"} \\
      Support vector machines & \CRANpkg{e1071} \citep{e1071-pkg} & \code{"svm"} \\
                              & \CRANpkg{kernlab} \citep{kernlab-pkg} & \code{"ksvm"} \\
      \bottomrule
  \end{tabular}
  \caption{Models specifically supported by the \pkg{package}.}
  \label{tab:models}
\end{table}
The \code{partial.default} method should be able to handle most other types of models not listed in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} or projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.

For illustration, we will use the (corrected) Boston housing data which are available from the \CRANpkg{mlbench} package \citep{mlbench-pkg}. These data contain the median value of owner-occupied homes in 506 U.S. census tracts in the Boston area, along with 13 independent variables such as the per capita crime rate by town. We begin by loading the data and omitting unimportant columns.
\begin{example}
data(BostonHousing2, package = "mlbench")  # mlbench must be installed!
boston <- BostonHousing2[, -c(1, 2, 5)]
\end{example}
Next, we fit a traditional random forest to the entire data set using the following code snippet:
\begin{example}
library(randomForest)
set.seed(101)  # for reproducibility
boston.rf <- randomForest(cmedv ~ ., data = boston, importance = TRUE)
varImpPlot(boston.rf)  # variable importance plot
\end{example}
The model fit is reasonable, with an \dfn{out-of-bag} $R^2$ of 0.89. The variable importance scores are displayed in Figure~\ref{fig:plotmo_vs_partial}. Both plots indicate that the percentage of lower status of the population (\code{lstat}) and the average number of rooms per dwelling (\code{rm}) are highly associated with the median value of owner-occupied homes. The question then arises, "What is the nature of these associations?" To help answer this, we can look at the partial dependence of \code{cmedv} on \code{lstat} and \code{rm}, both individually and together.

@@

As previously mentioned, the \code{randomForest} package has its own \code{partialPlot} function for displaying partial dependence of the response on a single predictor. For example, the following snippet of code plots the partial dependence of \code{cmedv} on \code{lstat}:
\begin{example}
partialPlot(boston.rf, pred.data = boston, x.var = "lstat")
\end{example}
The same plot (implemented in \pkg{lattice}) can be achieved using the \code{partial} function and setting \code{plot = TRUE}:
\begin{example}
library(partial)
partial(boston.rf, pred.var = "lstat", plot = TRUE)
\end{example}
For a more customizable plot, call \code{partial} with \code{plot = FALSE} and use the \code{plotPartial} function. \strong{Note:} the \dfn{pipe} operator provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is imported for writing more convenient code, as illustrated in snippet of code below.
\begin{example}
partial(boston.rf, pred.var = "lstat") %>%
  plotPartial(smooth = TRUE, lwd = 2, ylab = expression(f(lstat)))
\end{example}
The results are displayed in Figure~\ref{fig:pd_lstat} below.

@@

The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of model fits (not just random forests), (2) it will allow for any number of predictors to be used, and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section. For example, the following code chunk could be used to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv} based on a conditional random forest using various options in \code{plotPartial}. The results are displayed in Figure~\ref{fig:pd_lstat_rm}.
\begin{example}
library(party)
boston.crf <- cforest(cmedv ~ ., data = boston)
pd.lstat.rm <- partial(boston.rf, pred.var = c("lstat", "rm"))
plotPartial(pd.lstat.rm)
plotPartial(pd.lstat.rm, col.regions = colorRampPalette(c("white", "blue")))
plotPartial(pd.lstat.rm, contour = FALSE, zlab = "cmedv", drape = TRUE,
            colorkey = TRUE, screen = list(z = -20, x = -60))
\end{example}

>>>>

\abstract{
Complex nonparametric models (e.g., neural networks, random forests, and support vector machines) are often used in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g. multiple linear regression). Unfortunately, it can be challenging for the uninitiated to understand the results of such models and explain them to management. Partial dependence plots (PDPs) offer a simple solution. Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be easily understood. These plots are especially useful in explaining the output from complex black box models like support vector machines or neural networks or constructing simpler parsimonious models on newly obtained data.
}

@@

Predictor importance is an important task in any supervised learning problem. However, ranking variables is only part of the story and once a subset of "important" features is identified it is often necessary to assess the relationship between them (or subset thereof) and the response. This can be done in many ways, but in machine learning it is often accomplished by constructing \dfn{partial dependence plots} (PDPs); see \citet{friedman-2000-greedy} for details. PDPs help visualize the relationship between a subset (typically 1-3) of the predictors and the response while accounting for the average effect of the other independent variables in the model. They are particularly effective with black box models like random forest and support vector machines.
%PDPs offer a low-dimensional graphical rendering of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ for any type of fitted models and are especially useful in visualizing the relationships discovered by complex machine learning algorithms such as a \dfn{random forest}.

Let $\boldsymbol{x} = \left\{x_1, x_2, \dots, x_n\right\}$ represent the predictors in a model whose prediction function is $\widehat{f}\left(\boldsymbol{x}\right)$. If we partition $\boldsymbol{x}$ into an interest set, $\boldsymbol{z}_s$, and its compliment, $\boldsymbol{z}_{c} = \boldsymbol{x} \setminus \boldsymbol{z}_s$, then the partial dependence of the response on $\boldsymbol{z}_s$ is defined as
\begin{equation}
\label{eqn:pdf}
\bar{f}_s\left(\boldsymbol{z}_s\right) = \frac{1}{n}\sum_{i = 1}^n\widehat{f}\left(\boldsymbol{z}_s,\boldsymbol{z}_{i, c}\right),
\end{equation}
where $\boldsymbol{z}_{i, c}$ $\left(i = 1, 2, \dots, n\right)$ are the values of $\boldsymbol{z}_c$ that occur in the training sample; that is, weaverage out the effects of all the other predictors in the model. Let $x_1$ be the predictor variable of interest with unique values $\left\{x_{11}, x_{12}, \dots, x_{1k}\right\}$. The partial dependence of the response on $x_1$ can be constructed as follows:
\begin{enumerate}
  \item For $i \in \left\{1, 2, \dots, k\right\}$:
  \begin{enumerate}
    \item Create a copy of the training data and replace the original values of $x_1$ with the constant $x_{1i}$.
    \item Compute the vector of predicted values from the modified copy of the training data.
    \item Compute the average prediction to obtain $\bar{f}_1\left(x_{1i}\right)$.
  \end{enumerate}
  \item Plot the pairs $\left[x_{1i}, \bar{f}_1\left(x_{1i}\right)\right]$ for $i = 1, 2, \dotsc, k$.
\end{enumerate}
This can be quite computationally intensive since the algorithm involves a pass over the training records for each set of combinations of $\boldsymbol{z}_s$. Fortunately, this parallelized quite easily. This algorithm can also be easily extended to larger subsets of 2 or more features as well.

Limited implementations of Friedman's PDPs are available in packages \CRANpkg{randomForest} \citep{randomForest-pkg} and \CRANpkg{gbm}, among others; these are limited in the sense that they only apply to the models fit using these packages. While the \pkg{randomForest} implementation will only allow for a single predictor, the \pkg{gbm} implementation can deal with any subset of the predictor space; however, these implementations only apply to models fit using the respective package. Partial dependence functions are not restricted to tree-based models; they can be applied to any supervised learning algorithm (e.g., neural networks). However, to our knowledge, there is no general package for constructing PDPs  in R (e.g., PDPs for a \dfn{conditional random forest} as implemented by the \code{cforest} function in the \CRANpkg{party} and \CRANpkg{partykit} packages; see \citet{party-pkg} and \citet{partykit-pkg}, respectively). The \CRANpkg{partial} \citep{partial-pkg} package tries to close this gap by offering a general framework for PDPs that can be applied to several types of fitted models.

@@

Currently, only two functions are exported by \pkg{partial}:
\begin{itemize}
  \item \code{partial}
  \item \code{plotPartial}
\end{itemize}
The \code{partial} function evaluates the partial dependence \eqref{eqn:pdf} from a fitted model over a grid of predictor values. If \code{plot = FALSE} (the default), \code{partial} returns a data frame with an additional class: \code{"partial"}. The columns are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{y} and contains the partial dependence function values. If \code{plot = TRUE}, then \code{partial} calls \code{plotPartial} and returns a \code{"lattice"} object representing the PDP. For more advanced plotting, the \code{plotPartial} functions will take a \code{"partial"} object and display a more customizable lattice plot. Currently supported models are described in Table~\ref{tab:models}.
\begin{table}[htbp]
  \begin{tabular}{p{4cm}ll}
    \toprule
      Model & Package & Class \\
      \midrule
      Bagged decision trees     & \CRANpkg{ipred} \citep{ipred-pkg} & \code{"classbagg"}, \code{"regbagg"} \\
                                & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"bagging"} \\
      Boosted decision trees    & \pkg{gbm}      & \code{"gbm"} \\
                                & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"boosting"} \\
      Conditional random forest & \pkg{party}    & \code{"RandomForest"} \\
                                & \pkg{partykit} & \code{"cforest"} \\
      Decision trees            & \CRANpkg{rpart} \citep{rpart-pkg} & \code{"rpart"} \\
                                & \pkg{party}    & \code{"BinaryTree"} \\
                                & \pkg{partykit} & \code{"constparty"}/\code{"party"} \\
      Linear model              & \pkg{stats}    & \code{"lm"} \\
      Multivariate adaptive regression splines (MARS) & \CRANpkg{earth} \citep{earth-pkg} & \code{"earth"} \\
      Random forest & \pkg{randomForest} & \code{"randomForest"} \\
      Support vector machines & \CRANpkg{e1071} \citep{e1071-pkg} & \code{"svm"} \\
                              & \CRANpkg{kernlab} \citep{kernlab-pkg} & \code{"ksvm"} \\
      \bottomrule
  \end{tabular}
  \caption{Models specifically supported by the \pkg{package}.}
  \label{tab:models}
\end{table}
The \code{partial.default} method should be able to handle most other types of models not listed in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} or projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.

For illustration, we will use the (corrected) Boston housing data which are available from the \CRANpkg{mlbench} package \citep{mlbench-pkg}. These data contain the median value of owner-occupied homes in 506 U.S. census tracts in the Boston area, along with 13 independent variables such as the per capita crime rate by town. We begin by loading the data and omitting unimportant columns.
\begin{example}
data(BostonHousing2, package = "mlbench")  # mlbench must be installed!
boston <- BostonHousing2[, -c(1, 2, 5)]
\end{example}
Next, we fit a traditional random forest to the entire data set using the following code snippet:
\begin{example}
library(randomForest)
set.seed(101)  # for reproducibility
boston.rf <- randomForest(cmedv ~ ., data = boston, importance = TRUE)
varImpPlot(boston.rf)  # variable importance plot
\end{example}
The model fit is reasonable, with an \dfn{out-of-bag} $R^2$ of 0.89. The variable importance scores are displayed in Figure~\ref{fig:plotmo_vs_partial}. Both plots indicate that the percentage of lower status of the population (\code{lstat}) and the average number of rooms per dwelling (\code{rm}) are highly associated with the median value of owner-occupied homes (\code{cmedv}). The question then arises, "What is the nature of these associations?" To help answer this, we can look at the partial dependence of \code{cmedv} on \code{lstat} and \code{rm}, both individually and together.

@@

As previously mentioned, the \code{randomForest} package has its own \code{partialPlot} function for visualizing the partial dependence of the response on a single predictor. For example, the following snippet of code plots the partial dependence of \code{cmedv} on \code{lstat}:
\begin{example}
partialPlot(boston.rf, pred.data = boston, x.var = "lstat")
\end{example}
The same plot can be achieved using the \code{partial} function and setting \code{plot = TRUE} (see the left side of Figure~\ref{fig:pd_lstat}) as in the folling example:
\begin{example}
library(partial)
partial(boston.rf, pred.var = "lstat", plot = TRUE)
\end{example}
The only difference is that \pkg{partia} uses \CRANpkg{lattice} \citep{lattice-pkg} to produce all of its displays.

For a more customizable plot, call \code{partial} with \code{plot = FALSE} and use the \code{plotPartial} function. \strong{Note:} the \dfn{pipe} operator provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is imported for writing more convenient code, as illustrated in snippet of code below.
\begin{example}
partial(boston.rf, pred.var = "lstat") %>%
  plotPartial(smooth = TRUE, lwd = 2, ylab = expression(f(lstat)))
\end{example}
The result is displayed in the right side of Figure~\ref{fig:pd_lstat} below.

@@

The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of model fits (not just random forests), (2) it will allow for any number of predictors to be used, and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section. For example, the following code chunk could be used to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv} based on a conditional random forest using various options in \code{plotPartial}. The results are displayed in Figure~\ref{fig:pd_lstat_rm}.
\begin{example}
library(party)
boston.crf <- cforest(cmedv ~ ., data = boston)
pd.lstat.rm <- partial(boston.rf, pred.var = c("lstat", "rm"))
plotPartial(pd.lstat.rm)
plotPartial(pd.lstat.rm, col.regions = colorRampPalette(c("white", "blue")))
plotPartial(pd.lstat.rm, contour = FALSE, zlab = "cmedv", drape = TRUE,
            colorkey = TRUE, screen = list(z = -20, x = -60))
\end{example}
Note that the default color map for contour plots is the Matplotlib "viridis" color map provided through the \CRANpkg{viridis} package \citep{viridis-pkg}.
